---
title: "Peer Assesment 1"
author: "akrosalsky"
date: "08/11/2015"
output: html_document
---
### Introduction
It is now possible to collect a large amount of data about personal movement 
using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone 
Up. These type of devices are part of the “quantified self” movement – a group
of enthusiasts who take measurements about themselves regularly to improve their 
health, to find patterns in their behavior, or because they are tech geeks. But 
these data remain under-utilized both because the raw data are hard to obtain 
and there is a lack of statistical methods and software for processing and 
interpreting the data.

This assignment makes use of data from a personal activity monitoring device. 
This device collects data at 5 minute intervals through out the day. The data 
consists of two months of data from an anonymous individual collected during 
the months of October and November, 2012 and include the number of steps taken 
in 5 minute intervals each day.


### Data
The data for this assignment can be downloaded from the course web site:

Dataset: Activity monitoring data [52K]

The variables included in this dataset are:

*steps: Number of steps taking in a 5-minute interval (missing values are coded 
as NA)

*date: The date on which the measurement was taken in YYYY-MM-DD format

*interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a 
total of 17,568 observations in this dataset.

----------

------------

### Assignment

Load the data from the CSV file

```{r, echo=TRUE}
activity <- read.csv("activity.csv")
library(lattice)
```

---------

#### What is mean total number of steps taken per day?

Calculate total number of steps taken per day, then make a histogram of the 
total number of steps taken each day
```{r, echo=TRUE}
daily <- (tapply(activity$steps, activity$date, sum))

daily <- daily[!is.na(daily)]

histogram(~tapply(steps, date, sum),
        data = activity,
        col = 5,
        ylab = "Number of Days",
        xlab = "Total Steps",
        main = "Total Steps per Day",
        breaks = 20)
        
```

```{r, echo=TRUE}
median <- median(daily)

mean <- mean(daily)
```  

The mean number of steps taken daily is `r as.integer(mean)` steps and the 
median is `r median` steps.

---------------

#### What is the average daily activity pattern?

```{r, echo=TRUE}
xyplot((tapply(steps[!is.na(steps)], interval[!is.na(steps)], mean)) ~ interval,
        data=activity,
        type="l", 
        col = 1,
        xlab = "5-minute interval",
        ylab = "Steps",
        main = "Average Steps per 5-Minute Interval") 

time <- (tapply(activity$steps[!is.na(activity$steps)], activity$interval
                [!is.na(activity$steps)], mean))
max_steps <- max(time)
max_int <- names(time[time == max_steps])
```

The five-minute interval `r max_int` contains the maximum number of steps, on 
average, across the dataset. This interval contains an average of `r max_steps` 
steps.

-------------------

#### Imputing missing values

```{r, echo=TRUE}
missing <- length(activity[is.na(activity)])
```

There are `r missing` missing values in the dataset.

To fill in the missing values, the mean for the 5-minute interval will be used. 
This is being done because there are entire days with no values, so using the 
mean or median for each day would still result in some days with missing values.

```{r, echo=TRUE}
replace_missing <- function(activity){
        interval_means <- tapply(activity$steps[!is.na(activity$steps)],
                                 activity$interval[!is.na(activity$steps)],mean)
        
        na_index <- which(is.na(activity$steps) == TRUE)
        
        for (i in na_index){
                interval <- (names(interval_means)==activity$interval[i])
                activity$steps[i] <- as.numeric(interval_means[which(interval)])
        }
        activity
}

new_activity <- replace_missing(activity)

```

##### Histogram

```{r, echo=TRUE}
histogram(~tapply(steps, date, sum),
        data = new_activity,
        col = 7,
        ylab = "Number of Days",
        xlab = "Total Steps",
        main = "Total Steps per Day",
        breaks = 20)
```

```{r, echo=TRUE}
days <- (tapply(new_activity$steps, new_activity$date, sum))

median <- median(days)
mean <- mean(days)
```  

With the missing values filled in, the mean number of steps taken daily is 
`r as.integer(mean)` steps and the median is `r as.integer(median)` steps.

The mean number of steps remain the same because the 5-minute inteval median was
used for filling in the missing values so the mean was unaffected. The median is
only slightly different because the mean and median were so similar in the first
part of the assignment.

---------------------
####Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}
days <- sapply(as.Date(new_activity$date), weekdays)
weekend <- c("Saturday", "Sunday")
week <- NULL
for(i in days){
        if (i %in% weekend){
                week <- c(week, "weekend")
        }else{week <- c(week, "weekday")}
}

new_activity[2]<-("date"=week)
```

To determine if there is a difference between activity levels on weekdays and 
weekends, a new variable must be added to the data frame, separating the data 
into weekdays and weekends. 

```{r, echo=TRUE}
library(dplyr)
weekend <- filter(new_activity, new_activity$date == "weekend")
head(weekend)
weekend <- tapply(weekend$steps, weekend$interval, mean)

weekday <- filter(new_activity, new_activity$date == "weekday")
weekday <- tapply(weekday$steps, weekday$interval, mean)

week_data <- data.frame("mean" = c(weekend, weekday), 
                        "date" = c(rep("weekend", length(weekend)),
                                   rep("weekday", length(weekday))), 
                        "interval" = rep(unique(new_activity$interval),2))

xyplot(mean ~ interval | date,
        data = week_data,
        type = "l",
        col = 1,
        layout = c(1,2),
        ylab = "Number of steps",
        main = "Weekend and Weekday Activity Patterns")

```
